---
title: "Mandatory Jail Sentence Policy and Traffic Fatalities"
output:
  html_document:
    df_print: paged
    fig_caption: yes
    number_sections: yes
  pdf_document: default
---

Team ID: 9

Zhikuan Quan (Model Building); Daidai Zhang (Interpretation of model); Wenfeng Chang (Model Analysis); Jinghui Li (Exploratory Data Analysis)

Github repo: "http://github.com/zhikuanquan/STA-207.git"

```{r,echo=FALSE,include=FALSE}
### Input Data ###
library(AER)
library(plm)
library(stargazer)
library(tidyverse)
data("Fatalities")
# Missing Values
# since 1982-1988 is no, we assume it is "no" in missing place(Need some paper/law to ensure)
tna<-which(is.na(Fatalities),arr.ind= TRUE) # CA 1988 missing: jail, service
Fatalities[28,15]<-"no" 

### DATA of interest ###
DATA<-Fatalities %>%
  #filter(complete.cases(.)) %>%
  transmute(fr = 10000*((fatal/pop)), jail = as.factor(jail),state = as.factor(state), year = as.factor(year),beertax = beertax,emppop = emppop,miles = miles,income = income,drinkage = as.factor(floor(drinkage)), unemp=unemp, spirits = spirits,service = as.factor(service),breath=as.factor(breath),afatal=10000*(afatal/pop))

```

# 1. Introduction


# 2. Exploratory Data Analysis
This study used a balanced panel data for the 48 states in United States (excluding Alaska, Hawaii, and the District of Columbia) from 1982 to 1988, with 336 observations in total. The vehicl fatality rate per year of each state was calculated by dividing the number of vehicle fatalities by state population, then multiplying by 10000 to represent the average fatality every 10000 people. As shown in Figure 1, the average fatality rates over 7 years are quite different across the country. Rhode Island has the lowest fatality rate (1.11) while New Mexico has the highest (3.65).  
```{r, echo=FALSE, include=FALSE}
library(AER)
data("Fatalities")
dat = Fatalities
dat$y = dat$fatal/dat$pop * 10000
dat00 = aggregate(y~state, data = dat, mean)
fatal_state = data.frame(fatality = dat00$y, state = unique(dat$state)) 
library(usmap)
library(ggplot2)
p1 = plot_usmap(regions = 'state', data = fatal_state, values = "fatality",
     exclude = c('AK', 'HI'))+ scale_fill_continuous(high = "#132B43", low = "#56B1F7", 
                name = "Average fatality") + theme(legend.position = "right")
```

```{r, echo=FALSE}
p1
```

__Figure 1 Average fatality rates over 1982 through 1988 across United States__

Unemployment rate (unemp), spirits consumption (spirits), per capita income (income), tax on beer (beertax) and mandatory jail sentence for the first DUI conviction (jail) were shown as important factors affecting the vehicle fatality rate (Ruhm, 1996), so they were also considered in this study. As shown in Figure 2, the average per capita income across the country kept increasing from 1982 to 1988, while spirits consumption,unemployment rate and tax on beer decreased. Average fatality stayed rather stable over 7 years, and states having mandatory jail sentence seems to have higher average fatality than those not (Figure 3), which is surprising and will be analyzed in the following. There was one missing value of mandatory jail sentence status for California in 1988, which was replaced with "no" after searching the online information and double-checking with the local government. Per capita income was removed from the statistic model to avoid collinearity because it has a high correlation (-0.55) with unemployment rate, and is more correlated to other variables than unemployment rate (Figure 4).

```{r, echo=FALSE}
par(mfrow = c(2,2))
dat01 = aggregate(income~year,data = dat, mean)
plot(1982:1988, dat01$income, type = 'b', xlab = 'Year',
     ylab = 'Per capita income, $', main = 'Income')
dat02 = aggregate(spirits~year,data = dat, mean)
plot(1982:1988, dat02$spirits, type = 'b', xlab = 'Year',
     ylab = 'Per capita spirits consumption, gal', main = 'Spirits consumption')
dat03 = aggregate(unemp~year,data = dat, mean)
plot(1982:1988, dat03$unemp, type = 'b', xlab = 'Year',
     ylab = 'Unemployment rate, %', main = 'Unemployment rate')
dat04 = aggregate(beertax~year,data = dat, mean)
plot(1982:1988, dat04$beertax, type = 'b', xlab = 'Year',
     ylab = 'Tax on case of beer, $', main = 'Beer tax')
```

__Figure 2 The trends of income, spirits comsumption, unemployment rate and tax on beer from 1982 to 1988__

```{r, echo=FALSE}
ggplot(dat, aes(x = year, y = y, fill = jail)) + geom_boxplot() + labs(y = "Average fatality")
```

__Figure 3 The effect of mandatory jail sentence on average fatality from 1982 to 1988__

```{r, echo=FALSE}
dat1 = dat[lapply(dat, typeof)== 'double']
plot(dat1[,c(1,2,3,5)])
```

__Figure 4 Pairwise correlation among income, spirits comsumption, unemployment rate and tax on beer__



# 3. Fixed Effect Panel Model
## 3.1 Model Building
Fixed effect panel model is efficient since it controls for all time-invariant differences between the individuals. For instance, the transportation industry or the climate of specific state are fixed over short time, but they actually influence the fatality rate of traffic accident. In this case, fixed effect panel model is used to reduced the omitted variable biases of not accounting for these variables. 

Through exploratory data analysis, the US traffic fatalities data in 48 states except for Alaska and Hawaii from 1982 to 1988 is balanced. In addition, whether there is mandatory jail sentence in the state seems to affect the reduced traffic fatality rate in each state. Since the unemployment rate, tax on case of beer and spirits consumption are correlated to the traffic fatality rate of the year in each state, we consider them as the time-variant variables in our fixed effect panel model. In this case, we set the model notations as below:

- Response variable $Y_{it}$: the traffic fatality rate of state $i$ in time $t$. In this case, we have $Y_{it}=(10000\times F_{it})/P_{it}$, where $F_{it}$ represents the number of vehicle fatalities of the state $i$ in year $t$ and $P_{it}$ is the population of the state $i$ in year $t$. In this report, the traffic fatality rate can be interpreted as the number of death caused by traffic per 10000 population;

- Time fixed effect $\alpha_t$: time specific intercept. It represents any change over time (1982-1988) that affects all observational units in the same way. The level of time effect is 7;

- Mandatory jail sentence indicator $X_{it}$: $X_{it}=1$ when there is mandatory jail sentence of state $i$ in time $t$; $X_{it}=0$ when there is no mandatory jail sentence of state $i$ in time $t$;  

- Covariates vector $Z_{it}$: potential time-varying covariates. In this case, we mainly focus on three variables: the unemployment rate, beer tax and spirits consumption.

- State fixed effect $S_i$: In this case, each level of $S_i$ represents one of 48 states in US excluding Alaska and Hawaii.

The fixed effect model equation is:
$$Y_{it}=\alpha_t+\beta X_{it}+\gamma Z_{it}+S_i+\epsilon_{it}$$
where $\beta$ and vector $\gamma$ are regression coefficients and $\epsilon_{it}$ is error term. In fixed effect model, we assume that: (1) The normality of error terms; (2) The homogeneity of variance assumption; (3) Independent assumption: in time series data, the error terms are allowed to be autocorrelated within states, which means that there is autocorrelation relationship across time in one specific state. However, in a fixed time, the error terms should be independent across states.

## 3.2 Fitting result
Because our data set is from observation,in order to draw inferences about the causal effect of an active treatment versus a control treatment based on data in which the investigator did not decide which units would receive treatment and which would receive control. A useful way is to use logistic regression to estimate their propensity scores, so that we can estimate each unit’s probability of being treated as a function of covariates.

```{r,echo=FALSE}
library(MatchIt)
library(dplyr)
library(ggplot2)
Jail<-as.integer(ifelse(DATA$jail=="yes",1,0))
Data1<-data.frame(Jail,DATA[-2])
log1<-glm(Jail~unemp+beertax+spirits+year+state-1,data=Data1,family=binomial())
predf1<-data.frame(pr_score1=predict(log1,type="response"),Jail=log1$model$Jail)
summary(log1)
labs <- paste("probability:",c("jail","no jail"))
predf1 %>%
  mutate(Jail = ifelse(Jail == "1", labs[1], labs[2])) %>%
  ggplot(aes(x = pr_score1)) +
  geom_histogram(color = "white") +
  facet_wrap(~Jail) +
  xlab("Probability of jail") +
  theme_bw()

```


```{r,echo=FALSE}
DATA_nomiss <- Data1%>%na.omit()
mod_match <- matchit(Jail~unemp+beertax+spirits,method = "nearest", data = DATA_nomiss)
data_match<-match.data(mod_match)
dim(data_match)
```

```{r}
data_cov<-c("unemp","beertax","spirits")
Data1%>%
  group_by(Jail) %>%
  select(one_of(data_cov))%>%
  summarise_all(funs(mean))
data_cov<-c("unemp","beertax","spirits")
data_match %>%
  group_by(Jail) %>%
  select(one_of(data_cov))%>%
  summarise_all(funs(mean))
```

From the result, we can see that after matching,  we can attain a high degree of balance on the covariates included in the model.

```{r}
plm1<- plm(fr ~ Jail+unemp+beertax+spirits, data = data_match, index = c("state","year"),model = "within", effect = "twoways")
summary(plm1)
```
Using the data set after matching, we can find that jail is not significant.

##3.3 Model diagnostic
```{r}
lm1<- lm(fr ~ Jail+unemp+beertax+spirits+year+state-1, data = data_match)
plot(lm1)

```

```{r}
library(nortest)
```


```{r}
boxCox(lm(fr ~ Jail+unemp+beertax+spirits+year+state-1, data = data_match))
```

```{r}
lm2<-lm(log(fr) ~ Jail+unemp+beertax+spirits+year+state-1, data = data_match)
plot(lm2)
```

Check normality
```{r}
ad.test(lm2$residuals)$p.value
```

Check equal variance
```{r}
install.packages("lmtest")
```
```{r}
library(lmtest)
```

```{r}
bptest(lm2) 
```

```{r}
summary(lm2)
```








